---
title: "Seurat PBMC 3k tutorial using TENxPBMCData"
author: "Kevin Rue-Albrecht"
date: "27/11/2018"
output:
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=FALSE)
```

Fetch the SingleCellExperiment object using the `TENxPBMCData`.
The first time download from the web and cache locally; subsequently from the local cache.

```{r}
library(TENxPBMCData)
tenx_pbmc3k <- TENxPBMCData(dataset="pbmc3k")
tenx_pbmc3k
```

Prepare a sparse matrix that emulates the first section of the tutorial.

```{r}
library(Matrix)
pbmc.data <- as(counts(tenx_pbmc3k), "Matrix")
pbmc.data <- as(pbmc.data, "dgTMatrix")
colnames(pbmc.data) <- paste0("Cell", seq_len(ncol(pbmc.data)))
rownames(pbmc.data) <- make.unique(rowData(tenx_pbmc3k)[, "Symbol_TENx", drop=TRUE])
```

From here on, follow the [Seurat tutorial](https://satijalab.org/seurat/pbmc3k_tutorial_1_4.html) to the letter (code copied from https://satijalab.org/seurat/pbmc3k_tutorial.html on 2018-11-24).

```{r}
library(Seurat)
library(dplyr)
# Initialize the Seurat object with the raw (non-normalized data).  Keep all
# genes expressed in >= 3 cells (~0.1% of the data). Keep all cells with at
# least 200 detected genes
pbmc <- CreateSeuratObject(raw.data=pbmc.data, min.cells=3, min.genes=200, project="10X_PBMC")
```

```{r}
# The number of genes and UMIs (nGene and nUMI) are automatically calculated
# for every object by Seurat.  For non-UMI data, nUMI represents the sum of
# the non-normalized values within a cell We calculate the percentage of
# mitochondrial genes here and store it in percent.mito using AddMetaData.
# We use object@raw.data since this represents non-transformed and
# non-log-normalized counts The % of UMI mapping to MT-genes is a common
# scRNA-seq QC metric.
mito.genes <- grep(pattern="^MT-", x=rownames(x=pbmc@data), value=TRUE)
percent.mito <- Matrix::colSums(pbmc@raw.data[mito.genes, ])/Matrix::colSums(pbmc@raw.data)

# AddMetaData adds columns to object@meta.data, and is a great place to
# stash QC stats
pbmc <- AddMetaData(object=pbmc, metadata=percent.mito, col.name="percent.mito")
VlnPlot(object=pbmc, features.plot=c("nGene", "nUMI", "percent.mito"), nCol=3)
```

```{r}
# GenePlot is typically used to visualize gene-gene relationships, but can
# be used for anything calculated by the object, i.e. columns in
# object@meta.data, PC scores etc.  Since there is a rare subset of cells
# with an outlier level of high mitochondrial percentage and also low UMI
# content, we filter these as well
par(mfrow=c(1, 2))
GenePlot(object=pbmc, gene1="nUMI", gene2="percent.mito")
GenePlot(object=pbmc, gene1="nUMI", gene2="nGene")
```

```{r}
# We filter out cells that have unique gene counts over 2,500 or less than
# 200 Note that low.thresholds and high.thresholds are used to define a
# 'gate'.  -Inf and Inf should be used if you don't want a lower or upper
# threshold.
pbmc <- FilterCells(object=pbmc, subset.names=c("nGene", "percent.mito"), 
    low.thresholds=c(200, -Inf), high.thresholds=c(2500, 0.05))
```

```{r}
pbmc <- NormalizeData(object=pbmc, normalization.method="LogNormalize", 
    scale.factor=10000)
```

```{r}
pbmc <- FindVariableGenes(object=pbmc, mean.function=ExpMean, dispersion.function=LogVMR, 
    x.low.cutoff=0.0125, x.high.cutoff=3, y.cutoff=0.5)
```

```{r}
length(x=pbmc@var.genes)
```

```{r}
pbmc <- ScaleData(object=pbmc, vars.to.regress=c("nUMI", "percent.mito"))
```

```{r}
pbmc <- RunPCA(object=pbmc, pc.genes=pbmc@var.genes, do.print=TRUE, pcs.print=1:5, 
    genes.print=5)
```

```{r}
# Examine and visualize PCA results a few different ways
PrintPCA(object=pbmc, pcs.print=1:5, genes.print=5, use.full=FALSE)
```

```{r}
VizPCA(object=pbmc, pcs.use=1:2)
```

```{r}
PCAPlot(object=pbmc, dim.1=1, dim.2=2)
```

```{r}
# ProjectPCA scores each gene in the dataset (including genes not included
# in the PCA) based on their correlation with the calculated components.
# Though we don't use this further here, it can be used to identify markers
# that are strongly correlated with cellular heterogeneity, but may not have
# passed through variable gene selection.  The results of the projected PCA
# can be explored by setting use.full=T in the functions above
pbmc <- ProjectPCA(object=pbmc, do.print=FALSE)
```

```{r}
PCHeatmap(object=pbmc, pc.use=1, cells.use=500, do.balanced=TRUE, label.columns=FALSE)
```

```{r}
PCHeatmap(object=pbmc, pc.use=1:12, cells.use=500, do.balanced=TRUE, 
    label.columns=FALSE, use.full=FALSE)
```

Small deviation from the tutorial. Skip the lengthy JackStraw computation.

```{r}
# NOTE: This process can take a long time for big datasets, comment out for
# expediency.  More approximate techniques such as those implemented in
# PCElbowPlot() can be used to reduce computation time
if (FALSE) {
    pbmc <- JackStraw(object=pbmc, num.replicate=100, display.progress=FALSE)
    JackStrawPlot(object=pbmc, PCs=1:12)
}
```

```{r}
PCElbowPlot(object=pbmc)
```

```{r}
# save.SNN=T saves the SNN so that the clustering algorithm can be rerun
# using the same graph but with a different resolution value (see docs for
# full details)
pbmc <- FindClusters(object=pbmc, reduction.type="pca", dims.use=1:10, 
    resolution=0.6, print.output=0, save.SNN=TRUE)
```

```{r}
PrintFindClustersParams(object=pbmc)
```

```{r}
pbmc <- RunTSNE(object=pbmc, dims.use=1:10, do.fast=TRUE)
```

```{r}
# note that you can set do.label=T to help label individual clusters
TSNEPlot(object=pbmc)
```

Save the `seurat` object

```{r}
saveRDS(pbmc, file="pbmc3k_tutorial.rds")
```

Save the equivalent `SingleCellExperiment` object

```{r}
sce_pbmc <- as.SingleCellExperiment(pbmc)
saveRDS(sce_pbmc, file="pbmc3k_tutorial.sce.rds")
```

Note that the marker genes used to later assign cell identities to the various clusters are listed below

```{r}
## Markers and cell type annotations by cluster
# Cluster ID	Markers	        Cell Type
# 0             IL7R	        CD4 T cells
# 1             CD14, LYZ	    CD14+ Monocytes
# 2	            MS4A1           B cells
# 3	            CD8A	        CD8 T cells
# 4	            FCGR3A, MS4A7	FCGR3A+ Monocytes
# 5	            GNLY, NKG7	    NK cells
# 6	            FCER1A, CST3	Dendritic Cells
# 7	            PPBP	        Megakaryocytes
```
