---
title: "Applying Gene Signatures to the Seurat PBMC 3k Tutorial"
author: "Kevin Rue-Albrecht"
date: "`r BiocStyle::doc_date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=FALSE)
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(unisets)
    library(hancock)
    library(S4Vectors)
    library(knitr)
    library(ComplexHeatmap)
    library(ggplot2)
    library(cowplot)
})
```

# Example data set

First, load the `r Biocpkg("SingleCellExperiment")` object of [10X Genomics](www.10xgenomics.com) 2,700 peripheral blood mononuclear cells (PBMC) preprocessed [here](0-Seurat_PBMC3k_tutorial_TENxPBMCData.html).

```{r, }
sce <- readRDS("pbmc3k_tutorial.sce.rds")
sce
```

# Gene signatures

Declare the predefined gene signatures used in the [Seurat - Guided Clustering Tutorial](https://satijalab.org/seurat/pbmc3k_tutorial_1_4.html)

```{r}
geneLists <- list(
    "CD4 T cells"=c("IL7R"),
    "CD14+ Monocytes"=c("CD14", "LYZ"),
    "B cells"=c("MS4A1"),
    "CD8 T cells"=c("CD8A"),
    "FCGR3A+ Monocytes"=c("FCGR3A", "MS4A7"),
    "NK cells"=c("GNLY", "NKG7"),
    "Dendritic Cells"=c("FCER1A", "CST3"),
    "Megakaryocytes"=c("PPBP")
)
basesets <- as(geneLists, "BaseSets")
basesets
```

# Applying signatures to a data set

## The `predict` method

The `r Githubpkg("kevinrue/unisets")` `BaseSets` object can immediately be supplied to the `predict` S3 method, in combination with the `r Biocpkg("SummarizedExperiment")` object to annotate.

```{r}
sce <- predict(basesets, sce, method="ProportionPositive", cluster.col="ident")
```

## Prediction outputs

The updated `r Biocpkg("SummarizedExperiment")` object is returned, with a number of additional metadata.

The key output is the cell identity predicted for each cell in the object.
This information is stored in `colData(sce)[["hancock"]][["prediction"]]`, or `sce$hancock$prediction`, in short.

```{r}
table(sce$hancock$prediction)
```

With `method="ProportionPositive"`, all cells in each cluster are annotated with the predominant signature in their associated cluster.
As such, the cell type prediction may be tabulated against the cluster membership of each cell.

```{r}
table(sce$hancock$prediction, sce$ident)
```

This view can potentially highlight multiple clusters labelled by the same predicted label as a sign of overclustering.

In addition to the final predictions, the updated `r Biocpkg("SummarizedExperiment")` object also contain some additional information tracing to the process of prediction.
That information can be found in `metadata(sce)[["hancock"]]`.

```{r}
names(metadata(sce)[["hancock"]])
```

In particular:

- `GeneSets` stores the object that provided the gene set signatures used for the prediction
- `method` stores the name of the method used to make the predictions
- `packageVersion` stores the version of the `hancock` package used to make the predictions

Specifically, `method="ProportionPositive"` adds the following metadata items:

- `ProportionPositiveByCluster` stores the matrix indicating the proportion of cells in each cluster that are positive for each signature
- `TopSignatureByCluster` stores a named vector that indicates the predominant signature for each cluster. It is shown below.

```{r}
metadata(sce)[["hancock"]][["TopSignatureByCluster"]]
```

# Visualizing predictions

## Heat map

It is possible to visualize the proportion of cells positive for each signature in each cluster as a heat map.
Note how the diagonal (displaying the tutorial annotations) shows high proportions of cells matching the annotated signature.

```{r}
plotProportionPositive(sce, cluster_rows=FALSE, cluster_columns=FALSE)
```

We can also let signatures and clusters of cells cluster in the heat map.
Unsurprisingly, the clusters annotated as subtypes of monocytes show some similarity, as do the clusters annotated as T cell subtypes.

```{r}
plotProportionPositive(sce)
```

## Reduced dimension

With a bit more code, we can also display the expression of each marker on t-SNE result.

```{r, fig.height=10}
ggList <- list()
for (markerName in unique(unlist(elementIds(basesets)))) {
    ggDataFrame <- data.frame(
        reducedDim(sce, "TSNE"),
        logcounts=assay(sce, "logcounts")[markerName, ]
    )
    cellIdentity <- subset(as.data.frame(basesets), element == markerName, "set", drop=TRUE)
    gg <- ggplot(ggDataFrame, aes(tSNE_1, tSNE_2, color=logcounts)) +
        geom_point(size=0.1) +
        scale_color_viridis_c() +
        labs(title=markerName, subtitle=cellIdentity, x=NULL, y=NULL) +
        theme(
            axis.text = element_blank(), axis.ticks = element_blank(),
            text = element_text(size=rel(3)), legend.text = element_text(size=rel(2)),
            legend.position = "bottom")
    ggList[[markerName]] <- gg
}
cowplot::plot_grid(plotlist = ggList)
```

# See also

- [Learning Gene Signatures from the Seurat PBMC 3k Tutorial](2-learn-signatures.html)

<button onclick="topFunction()" id="myBtn" title="Go to top">Back to top</button>
