---
title: "Learn Signatures from the PBMC 3k data set"
author: "Kevin Rue-Albrecht"
date: "19/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=FALSE)
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Hancock)
    library(GeneSet)
})
```

Load the preprocessed SingleCellExperiment object of 10x PBMC 3k.

```{r, }
sce <- readRDS("pbmc3k_tutorial.sce.rds")
sce
```

According to the tutorial, cluster 0 comprises CD4 T cells marked by the expression of IL7R.
When applying the IL7R signature, we have shown that indeed IL7R is expressed at a relatively high frequency in cluster 0 and much less frequently detected in all other clusters.

Now, let us programmatically identify markers by asking the question:
"Which genes are detected at a high frequency in cluster 0, but rarely detected in all other clusters?"

```{r}
tgs <- learnSignature(sce, cluster.col="ident")
knitr::kable(tgs)
```

Note that the table above does highlight some of the markers suggested in the [Seurat tutorial](https://satijalab.org/seurat/pbmc3k_tutorial_1_4.html).
Namely, CD14 (CD14+ Monocytes), MS4A1 (B cells), CD8A (CD8 T cells), FCER1A (Dendritic Cells), PPBP (Megakaryocytes).

```{r}
seuratTutorialMarkers <- tbl_geneset(
    "CD4 T cells"=c("IL7R"),
    "CD14+ Monocytes"=c("CD14", "LYZ"),
    "B cells"=c("MS4A1"),
    "CD8 T cells"=c("CD8A"),
    "FCGR3A+ Monocytes"=c("FCGR3A", "MS4A7"),
    "NK cells"=c("GNLY", "NKG7"),
    "Dendritic Cells"=c("FCER1A", "CST3"),
    "Megakaryocytes"=c("PPBP")
)
seuratTutorialMarkers
```

Let us visualize the top 2 markers for each cluster.
The cell type identity associated with markers mentioned in the [Seurat tutorial](https://satijalab.org/seurat/pbmc3k_tutorial_1_4.html) is indicated in each panel subtitle. 

```{r, fig.height=12}
require(ggplot2)
ggList <- list()
for (markerName in tgs$gene) {
    ggDataFrame <- data.frame(
        reducedDim(sce, "TSNE"),
        logcounts=assay(sce, "logcounts")[markerName, ]
    )
    seuratIdentity <- as.character(subset(seuratTutorialMarkers, gene == markerName, "set", drop=TRUE))
    gg <- ggplot(ggDataFrame, aes(tSNE_1, tSNE_2, color=logcounts)) +
        geom_point(size=0.1) +
        scale_color_viridis_c() +
        labs(title=markerName, subtitle=seuratIdentity, x=NULL, y=NULL) +
        theme(
            axis.text = element_blank(), axis.ticks = element_blank(),
            text = element_text(size=rel(3)), legend.text = element_text(size=rel(2)),
            legend.position = "bottom")
    ggList[[markerName]] <- gg
}
cowplot::plot_grid(plotlist = ggList)
```

