---
title: "Applying the xCellData Signatures to positive markers"
author: "Kevin Rue-Albrecht"
date: "`r BiocStyle::doc_date()`"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=FALSE)
suppressPackageStartupMessages({
    library(BiocStyle)
    library(knitr)
})
```

# Example data set

First, load the `r Biocpkg("SingleCellExperiment")` object of [10X Genomics](www.10xgenomics.com) 2,700 peripheral blood mononuclear cells (PBMC) preprocessed in a [previous vignette](0-Seurat_PBMC3k_tutorial_TENxPBMCData.html).

```{r, message=FALSE}
library(SingleCellExperiment)
sce <- readRDS("pbmc3k_tutorial.sce.rds")
sce
```

# Run paired t-tests to identify markers

First, we normalize and log-transform the count data.

```{r, message=FALSE}
library(scran)
library(scater)
sce <- computeSumFactors(sce)
sce <- normalize(sce)
sce
```

In this example, we look for _positive_ markers.
Namely, up-regulated genes in each cluster relative to all other clusters.

```{r, message=FALSE}
library(scran)
out <- findMarkers(sce, clusters=sce$seurat.ident, direction="up", lfc=1)
out
```

The result `out` is a list of `DataFrame`.
For each cluster, the log fold-change against every other cluster is reported, along with a combined p-value and FDR.

```{r}
colnames(out[[1]])
```

# Obtain gene signatures for relevant cell types

## Load xCellData

This package contains published gene signatures for a number of cell types.

```{r, message=FALSE}
library(xCellData)
library(unisets)
xcell <- xCellData()
xcell
```

## Prepare annotations for the analysis

In this vignette, we are about to use the `r Biocpkg("clusterProfiler")` package to compute the enrichment of published cell type signatures among sets of genes up-regulated in each cluster.
To do so, we need to match the [ensembl](https://www.ensembl.org/index.html) gene identifiers used in the PBMC data set to the [gene symbols](https://www.genenames.org) used in the published cell type gene signatures.

First, we use the `r Biocpkg("org.Hs.eg.db")` to convert the `r Githubpkg("kevinrue/xCellData")` published signatures from gene symbols to [ensembl](https://www.ensembl.org/index.html) gene identifiers.

```{r, message=FALSE}
library(org.Hs.eg.db)
TERM2GENE <- as.data.frame(xcell)[, c("set", "element")]
TERM2GENE$element <- mapIds(org.Hs.eg.db, TERM2GENE$element, "ENSEMBL", "SYMBOL")
```

The `r Biocpkg("clusterProfiler")` also requires a table mapping term identifiers to their name.
In this example, each signature identifier is already a descriptive name, so we provide empty strings.

```{r}
TERM2NAME <- data.frame(
    id = ids(setData(xcell)),
    name = rep("", nSets(xcell))
)
```

# Run the analysis

In this example, we apply Gene Set Enrichment Analysis (GSEA) on the lists of genes ranked by decreasing detection average log-fold change between each cluster and all other clusters.
Below, we display the 2 most signicantly enriched cell type gene signature for each cluster.
Note that for some clusters, none of the signatures reached signifcance.

```{r, message=FALSE, warning=FALSE}
library(clusterProfiler)
runGSEA <- function(clusterName, threshold=0, top=10) {
    x <- out[[clusterName]]
    x_lfc_colnames <- grep("^logFC\\.", colnames(x))
    x_rowmeans <- rowMeans(as.matrix(x[, x_lfc_colnames]))
    x_rowmeans <- sort(x_rowmeans, decreasing = TRUE)
    x_rowmeans <- x_rowmeans[x_rowmeans > threshold]
    y = GSEA(x_rowmeans, TERM2GENE=TERM2GENE, TERM2NAME=TERM2NAME, pvalueCutoff=1, verbose = FALSE)
    y <- as.data.frame(y)
    y <- head(y, top)
    y <- cbind(cluster = clusterName, y)
    y
}
resList <- lapply(levels(sce$seurat.ident), runGSEA, threshold=0, top=2)
res <- do.call(rbind, resList)
res <- as.data.frame(res)
show_columns <- setdiff(colnames(res), "core_enrichment")
kable(res[, show_columns])
```

# See also

- [Applying Gene Signatures to the Seurat PBMC 3k Tutorial](1-proportion_signature.html)
- [Learning Gene Signatures from the Seurat PBMC 3k Tutorial](2-learn-signatures.html)

<button onclick="topFunction()" id="myBtn" title="Go to top">Back to top</button>
